---
title: "An Introduction to the Reproducible Climate Futures package"
author: "Janelle Christensen"
date: "8-20-2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An Introduction to the Reproducible Climate Futures package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## About the rcf package

This package aims to make acquiring and working with CMIP5 [MACA v2-METDATA](http://www.climatologylab.org/maca.html) downscaled climate data faster and easier and to provide a number of summary statistics that can be used to visualize different climate futures. Ultimately, having access to this data and tools to work with it supports location-specific planning efforts that incorporate climate change.

The Global Circulation Models (GCM's) used within this package are the same 20 GCM's used in [MACA Version 2](http://www.climatologylab.org/maca.html). These models provide climate variables for historical and future models under both RCP4.5 and RCP8.5.

## Package goals

1. Given the user's area of interest, the package will facilitate the assignment of GCM's into four categories representative of plausible and divergent climate futures:
  (1) Warm & Wet
  (2) Warm & Dry (or Damp)*
  (3) Hot & Wet
  (4) Hot & Dry (or Damp)*

2. The user can then calculate threshold exceedances for 25 predefined climate metrics using one of three methods:
  (1) Quadrant
  (2) Corner
  (3) PCA
  
and one of three time periods:  
  (1) month  
  (2) season  
  (3) year
  

## About this tutorial

This vignette is the first in a series of two, and is intended for general users of the `rcf` package. Advanced users interested in Principal Component Analyses (PCA) may continue on with [the second vignette in the series](https://github.com/nationalparkservice/rcf/blob/main/vignettes/pca_intro.Rmd) 

At the end of this tutorial, the user should understand:

- the workflow of the `rcf` package
- naming conventions of output .csv files 
- different ways of temporally summarizing data (i.e., month, season, year)
- different methods for summarizing data in this package (i.e., quadrant, corner, PCA) and how to apply them


#### Expected programming knowledge

This vignette assumes familiarity with routine functions in the `tidyverse`, specifically `readr` and `read_csv()`; `filter()`, `mutate()` and `select()` in the package `dplyr`; and basic syntax for plot creation in `ggplot2`.


#### Important requirements and considerations

The files in this package, especially the raw data, are large. If the user is downloading data within this package using the included function `rcf_data()`, at least *500 MB* of space must be available. If data have been obtained from elsewhere, specifically for the `cf_pca()` function, the user must ensure that variable names in the `rcf` package match the column names of the outside dataset exactly.

This vignette uses a project directory in keeping with [best practices in R](https://rstats.wtf/project-oriented-workflow.html#work-in-a-project). As such, this vignette will walk through the steps as if this project is using a .Rproj file inside of a directory that will hold all information and resultant data for this project. This vignette will also use the `here()` package, rather than relative file paths.

## Get Started

#### Installation

Until approval on CRAN, you should download the development version of
`rcf` from [GitHub](https://github.com/) as follows:

``` r
# install.packages("devtools")
devtools::install_github("nationalparkservice/rcf")
```
 
The released version will soon be available on [CRAN](https://CRAN.R-project.org) 



## Package run-through 

#### Workspace setup

First, load dependencies.

```{r setup, message=FALSE, warning=FALSE}
library(rcf)
library(dplyr)
library(readr)
library(here)
library(ggplot2)
library(ggrepel)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#hidden because they are only affecting the behind the scenes of the tutorial
library(kableExtra)
library(devtools)
```

Then set the workspace that we will be using for all outputs from this package.

```{r}
my_directory <- "~"
```

This will set your directory to wherever the .Rproj file is located. Alternatively, you could create a folder in your project directory in which to save your files.

## 1. Download the data

For the purpose of this tutorial, data will be downloaded from the National Park Service via the `readr::read_csv()` function in the code below. Results will appear in your environment as a dataframe called `raw_data`. 

```{r}
raw_data <- read_csv("https://irmadev.nps.gov/DataStore/DownloadFile/660685")
```


Typically, data will be downloaded using the `rcf_data()` function from the `rcfdata` package, which is not currently available but will be included in a future release. `rcf_data()` downloads all data using the [`cft`](https://www.earthdatascience.org/cft/) package for all years from 1950 - 2099 for all 40 GCM-RCP combinations. Because of the amount of data being downloaded, this function takes about 4 hours to run. 

Variable units may be either "imperial" (the default) or "metric." Keep in mind that if using metric, variables that are calculated later, such as the heat index, can only be calculated with Fahrenheit. Results for those variables will convert the temperature to Fahrenheit first.

```{r, eval=FALSE}
# devtools::install_github("nationalparkservice/rcfdata")
# library(rcfdata)
# raw_data <- rcf_data(SiteID = "BAND", # National Park unit code (BAND = 'Bandelier National Park')
#                      latitude = 35.75758546, # Latitude of park centroid
#                      longitude = -106.3054344, # Longitude of park centroid
#                      units = "imperial")
```


The output of the `rcf_data()` function is a file named "BAND.csv", which will be saved in your working directory.  It includes daily values of:

```{r, echo=FALSE}

names_table <- data.frame(name = c("Minimum temperature", "Maximum temperature", "Precipitation", "Minimum relative humidity", "Maximum relative humidity", "Date", "Year"), df_name = c("tmin", "tmax", "precip", "rhmin", "rhmax", "date", "yr"))

names_table %>% 
  kable(col.names = c("Variable Name",
                      "Column Name")) %>% 
# Column names - can't change row names in kable.
  kable_styling(bootstrap_options = "striped",
                full_width = F,
                position = "center") 
# Full width (F), centered table.

```

The functions in this package are compatible with any climate data that meets the following requirements:

- Includes at least gcm, date, year, minimum temperature, maximum temperature, average temperature, and precipitation.*
- Provides daily observations
- Column names match the names in the table above (or are formatted accordingly after download)
- Has a range of at least 30 years of historical (modeled) data and 30 years of future modeled data

An example of an external data source that provides appropriate data can be [found here](https://climate.northwestknowledge.net/MACA/gallery_data.php)

*NOTE: data that downloads from the `rcf_data()` function or the `cft` package will include relative humidity, but the functions in this package will still work if your data does not include that variable*


## 2. Calculate thresholds

Now that you have your raw data, the next step in the package workflow is to calculate the threshold values for your site of interest. Threshold values are numeric values representing extreme conditions for a given metric (e.g., heat index) specific to the site of interest. This package evaluates threshold values for 25 metrics; these are listed in detail in the `Data` table in the [README file](https://github.com/nationalparkservice/rcf/blob/main/README.md) 

We calculate thresholds using the `calc_thresholds()` function.


```{r}
thresholds <- calc_thresholds("BAND", data = raw_data, directory = my_directory, units = "imperial")

glimpse(thresholds)

```

As you can see, the majority of the results from this function are `TRUE` or `FALSE` observations. This is because these are tests of whether or not a value on that day crossed over a specified threshold. `TRUE` means that the value is more extreme than the threshold; `FALSE` means it is not.

The output of this function is a csv file named "BAND_thresholds.csv"

For most users of this package, there will only be one more function to run before moving on to plotting. Advanced users can continue to follow along with this vignette or choose to move onto the [PCA intro](https://github.com/Janelle88/rcf_addendum/blob/master/RCF%20instructions%20-%20pca.pdf) vignette.

## 3. Calculate Climate Futures

The `cf_quadrant()` function allows users to calculate climate futures one of two ways:

1. *Quadrant method* - this method assigns to each model one of five climate futures (Warm Wet, Warm Dry (Damp)*, Hot Wet, Hot Dry (Damp), or Central), depending on change in precipitation and temperature from historic values and centered around one user-specified year in the future (e.g., 2040). The final values in the dataframe are the mean of all observations from all models in the quadrant, summarized by month, season or year, depending on the summarization method chosen.

2. *Corner method* - this method assigns one of four climate futures to the most extreme model in each quadrant. The final values in this dataframe are the mean of all observations in that model, summarized by month, season or year, depending on the summarization method chosen.

In total, there are 6 ways you can choose to summarize the threshold data using the `cf_quadrant()` function, but here we will use the corner method and summarize by year.

*GCMs will be labeled as "damp" rather than "dry" if the mean change in future precipitation for all quadrants and relative to historical values is greater than zero.*

```{r}
corner_year <- cf_quadrant("BAND", data = thresholds, future_year = 2040, summarize_by = "year", method = "corner", directory = my_directory)
```

The results from this function are two csvs:

1. "BAND_year_summary_c.csv" - the name of this output will change based upon what you enter into the function. Here, this file is named "year_summary" because we chose to summarize by year, but if we had chosen to summarize by month or season, this would be reflected in the file name. The "c" at the end is because we chose the corner method, and would have been labeled "q" if we had chosen the quadrant method.  

2. "BAND_future_means.csv" - this file assigns a quadrant to each model and identifies the most extreme model in each (if we have chosen the corner method). This csv will have to be read back into R.

Let's look at each file.

```{r}
glimpse(corner_year)
```

Because there are 2 csv's that come out of this function, the csv for the future means must be read in manually. 

```{r}
means <- read_csv("~/BAND_future_means.csv")

glimpse(means)
```

With these two files, we can now move on to plotting.


## 4. Plotting

Let's start with two plots that will show us where the models fall in relation to each other based on changes in temperature and precipitation, centered around our chosen year of 2040. 

First, we produce a graph that shows all models by color in their respective quadrants:


```{r, fig.width=15, fig.height=9, out.width="110%"}
ggplot(means, aes(x = tavg_change,
                  y = precip_change,
                  xmin = min(tavg_change) - 0.15 * min(tavg_change),
                  xmax = max(tavg_change) + 0.15 * max(tavg_change),
                  ymin = min(precip_change) - 0.15 * min(precip_change),
                  ymax = max(precip_change) + 0.15 * max(precip_change))) +
  geom_text_repel(aes(label = gcm,
                      color = cf),
                  position = position_jitter(0,.2)) + 
  geom_point(size=5,colour="black") +
  geom_point(aes(color = cf),
             size = 4) +
  labs(title = "Changes in climate means centered on 2040 by GCM", 
       x = "Change in temperature (F)", 
       y = "Change in precipitation (in)") + 
  scale_color_manual(values = c("gray", "#E10720",  "#8386CC", "darksalmon", "#12045C")) +
  scale_fill_manual(values = c("gray", "#E10720",  "#8386CC", "#darksalmon", "#12045C")) + 
  theme_minimal() +
  theme(text = element_text(size = 20))
```

Second, because we have chosen the corner method, let's make a graph that highlights the selected models:

```{r, fig.width=15, fig.height=9, out.width="110%"}
ggplot(means, aes(x = tavg_change,
                  y = precip_change,
                  xmin = min(tavg_change) - 0.15 * min(tavg_change),
                  xmax = max(tavg_change) + 0.15 * max(tavg_change),
                  ymin = min(precip_change) - 0.15 * min(precip_change),
                  ymax = max(precip_change) + 0.15 * max(precip_change))) +
  geom_text_repel(aes(label = gcm,
                      color = corner),
                  position = position_jitter(0,.2)) + 
  geom_point(aes(label = gcm,
                 color = corner),
             size = 5) +
  labs(title = "Changes in climate means centered on 2040 by GCM", 
       x = "Change in temperature (F)", 
       y = "Change in precipitation (in)") + 
  scale_color_manual(values = c("#E10720",  "#12045C", "darksalmon", "#8386CC")) +
  scale_fill_manual(values = c("#E10720",  "#12045C", "darksalmon", "#8386CC")) +
  theme_minimal() +
  theme(text = element_text(size = 20))
```

Now that we have seen where our models lie in relation to each other, let's make some plots using the threshold data.


First, let's see how the yearly precipitation changes over time in the four climate futures:

```{r, fig.width=15, fig.height=9, out.width="110%"}
# corner_year$time <- factor(corner_year$time, levels = c("Historical", "Future"))

ggplot(data = corner_year, aes(x = time, y = precip_yearly)) +
  geom_boxplot(aes(color = time,
                   fill = time),
               alpha = 0.2) +
  geom_jitter(aes(color = time),
              size = 2.5,
              alpha =.6) +
  facet_wrap(~corner) + #facet_grid for all next to each other
  scale_color_manual(values = c("#8386CC", "#12045C")) +
  scale_fill_manual(values = c("#8386CC", "#12045C")) +
  labs(y = "Yearly Precipitation (in)",
       title = "Yearly precipitation for individual GCMs representative of 4 climate futures") +
  theme_minimal() +
  theme(text = element_text(size = 20),
        legend.title = element_blank(),
        axis.title.x = element_blank())
```

This graph could be made for any of the variables in the output from the `cf_quadrant()` function, so let's look at a few more. 

Let's see how the number of days that exceed the historical 99th percentile of heat compare between the past and future:

```{r, fig.width=15, fig.height=9, out.width="110%"}
ggplot(data = corner_year, aes(x = time, y = temp_over_99_pctl)) +
  geom_boxplot(aes(color = time,
                   fill = time),
               alpha = 0.2) +
  geom_jitter(aes(color = time),
              size = 2.5,
              alpha =.6) +
  facet_wrap(~corner) +
  scale_color_manual(values = c("darksalmon", "#E10720")) +
  scale_fill_manual(values = c("darksalmon", "#E10720")) +
  labs(y = "Number of days per year",
       title = "Days where temperature exceeds the 99th percentile for\n individual GCMs representative of 4 climate futures") +
  theme_minimal() +
  theme(text = element_text(size = 20),
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5))
```


What about the growing season length between models? How does that compare between the four climate futures?

```{r, fig.width=15, fig.height=9, warning=FALSE, out.width="110%"}
corner_year_future <- corner_year %>%
filter(time %in% c("Future"))

ggplot(data = corner_year_future, aes(x = time, y = grow_length)) +
  geom_boxplot(aes(color = corner,
                   fill = corner),
               alpha = 0.2) +
  geom_jitter(aes(color = corner),
              size = 2.5,
              alpha =.6) +
  facet_grid(~corner) +
  scale_color_manual(values = c("#8FD834", "#72CC50", "#019875", "#00AEAD")) +
  scale_fill_manual(values = c("#8FD834", "#72CC50", "#019875", "#00AEAD")) +
  labs(y = "Length of growing season",
       title = "Growing season length for individual GCMs representative of 4 climate futures") +
  theme_minimal() +
  theme(text = element_text(size = 20),
        legend.title = element_blank(),
        axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank())
```

What if we want to compare by season or month? We can use the `cf_quadrant()` function again, this time summarizing by either month or season.

```{r}
corner_month <- cf_quadrant("BAND", data = thresholds, future_year = 2040, summarize_by = "month", method = "corner", directory = my_directory)
```

We can then graph by month or season by climate future, or across all climate futures.

```{r, fig.width=15, fig.height=9, warning=FALSE, out.width="110%"}
corner_month_hw <- corner_month %>% 
  filter(corner %in% c("Hot Wet")) %>% 
  mutate(month = lubridate::month(month, label = TRUE)) 

ggplot(data = corner_month_hw, aes(x = month, y = precip_monthly)) +
  geom_col(aes(fill = time,
               color = time), 
           position = "dodge",
           alpha = 0.7) +
  scale_color_manual(values = c("#8386CC", "#12045C")) +
  scale_fill_manual(values = c("#8386CC", "#12045C")) +
  theme_minimal() +
  theme(text = element_text(size = 20),
        legend.title = element_blank(),
        #axis.title.x = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
        #axis.text.x = element_blank()) +
  labs(y = "Monthly Precipitation (in)", x = " ",
        plot.title = element_text(hjust = 0.5)) +
  labs(y = "Monthly Precipitation (in)", x = " ",
       title = "Comparison of historical and future precipitation in the 'Hot Wet' climate future")
```

#### 5. Advanced Users

For users that want to try a more advanced approach to finding climate futures, the [Introduction to using PCA for model selection](https://github.com/Janelle88/rcf_addendum/blob/master/RCF%20instructions%20-%20pca.pdf) tutorial will walk you through the process.
